\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-07 k-planar.tex
A big change of notation

\newcommand{\PP}{\mathcal{P}}
\renewcommand{\SS}{\mathcal{S}}

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs.
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}), and \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `induced-universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$. This has recently been extended by \citet{EJM}, who show the existence of a `subgraph-universal' graph with $(1+\epsilon)n$ vertices that contains every $n$-vertex planar graph as a subgraph as well as the existence of an `induced universal' graph with $n^{1+o(1)}$ vertices \emph{and} edges.  The former result is the first progress on subgraph-universal graphs for planar graphs since the $O(n^{3/2})$-vertex construction of \citet{babai.chung.ea:on}.
% aking the first progress on this problem since the work of \citet{babai.xx} in 1984.
\end{compactitem}

\note{DW}{Delete `from 1992' and  `from 2007'? replace ``going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}'' by
	``starting with the work of \citet{kannan.naor.ea:implicit}''}
\note{PM}{Was this a referee request, or just because we don't need to sell it anymore?  I'm fine with the change either way, but didn't implement it.}

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is a subgraph of $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this definition.}. These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}, and \citet{DEMWW22} do so for graphs in any minor-closed class and with bounded maximum degree. See \cref{Generalisations} for more precise statements and see \citep{DHJLW21} for a survey on this topic.

@@

\note{PM}{Kind of, but I suggest we keep that paragraph and section anyway. Justification: They're both short and they describe the exact bounds for bounded genus graphs, which you don't get from the one-paragraph description in Section~1.0.  A}

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarProduct}, we actually prove a substantially stronger result than \cref{kPlanarProduct} using the following definition. A non-empty set $\mathcal{P}$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.  A path is \defin{trivial} if it has length 0 and \defin{non-trivial} otherwise.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem}
Each path $P\in\mathcal{P}$ is called a \defin{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \defin{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is a subgraph of $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$.
\end{obs}

The proof of \cref{AddDummy} is trivial: Given a $k$-plane embedding of a graph $G'$, create a planar graph $G$ by adding a dummy vertex at each crossing point. For each edge $vw\in E(G')$ there is a path $P$ in $G$ between $v$ and $w$ of length at most $k+1$ in which every internal vertex is a dummy vertex. Let $\PP$ be the set of such paths $P$. For each vertex $v$ of $G$, at most two paths in $\PP$ use $v$ as an internal vertex (since no original vertex of $G'$ is an internal vertex of a path in $\PP$). Thus $\PP$ is a $(k+1,2)$-shortcut system for $G$, such that $G'\subseteq G^\PP$. This idea can be pushed further to obtain a rough characterisation of $k$-planar graphs, which is interesting in its own right, and is useful for showing that various classes of graphs are $k$-planar (see \cref{Characterisation}).

@@

\begin{thm}
\label{ShortcutProduct}
Let $G$ be a subgraph of $H\boxtimes P \boxtimes K_\ell$, for some graph $H$ of treewidth at most $t$ and for some path $P$.
Let $\PP$ be a $(k,d)$-shortcut system for $G$. Then $G^\PP$ is a subgraph of $J\boxtimes P\boxtimes K_{d\ell(k^3+3k)}$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$ and some path $P$.
\end{thm}

@@

\note{DW}{Use $\SS$ for shortcut system, okay?}
\note{PM}{Ok.}

@@

\cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Here is one example. The \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} then imply:

@@

%The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
%
%\begin{thm}
%Let $\PP$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\PP$ is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\PP$ for $G$, $G^\PP\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\PP$ for $G$, $G^\PP \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\note{DW}{delete the above three theorems, they are never used}

@@

The following result is the main technical contribution of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so too does $G^\PP$ for every shortcut system $\PP$ of $G$.

\begin{restatable}{thm}{mmg}
	\label{ShortcutPartition}
	Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$.
\end{restatable}

@@

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\mathcal{P}$ of a graph $G$, a $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\mathcal{P}}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\PP$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\mathcal{P}}$ but eliminates special cases in some of our proofs.

@@

For a node $x\in V(T)$, we say that a shortcut $P\in\mathcal{P}$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\mathcal{P}$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

@@

\begin{proof}
  Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\mathcal{P}, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\PP$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$.

@@

\note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\mathcal{S}$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\mathcal{S}$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $J$.)

\referee{2}{Page 8.  Can you add a brief explanation why $\mathcal{S}$ is a
	partition of $V(G)$?  It is clear that the sets in $\mathcal{S}$ are
	disjoint, but why do they cover $V(G)$?}
\note{PM}{I added an explanation in a footnote since this is obvious.}

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\mathcal{S}$ has small layered width with respect to the layering $\mathcal{L}$ and that
  \item $J$ has small treewidth.
\end{compactenum}

@@

First we need to understand the relationship between sets $S_x\in\mathcal{S}$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$, with $v\in V(P)$, that crosses $x$.  The path $P$ contains a proper subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

Next we complete Step~(i) and show that $\mathcal{S}$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

\begin{proof}
  Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v$ participates in $x$.
  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes at most one vertex, itself, to $S_x\cap L_i$.

@@

  Otherwise, consider some path $P\in\mathcal{P}$ that contains $w$ as an internal vertex.  If $w\in L_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap L_i$.  If $w\in L_{i-1}\cup L_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap L_i$. If $w\in L_{i-j}\cup L_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap L_i$.

  For any $j$, the number of vertices $w\in L_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore,
  \[  |S_x\cap L_i|\le d\ell  \, \Big(k+1 + 2k + \sum_{j=2}^k 2(k-j)\Big) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +3) \enspace . \qedhere
  \]
\end{proof}

@@

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(J)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\mathcal{P}}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\mathcal{P}$ contains a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

  Since $w\in S_{\delta}$, $a(w)=x_\delta$, so $w$ participates in $x_\delta$, so at least one of the following is true:
  \begin{compactenum}
    \item There exists $w'\in V(G)$ such that $\mathcal{P}$ contains a $ww'$-shortcut $P_{ww'}$ that has an internal vertex in $Y_{\delta}$; or
    \item $w\in Y_\delta$; in this case, define $w':=w$ and let $P_{ww'}$ be the 1-vertex path $w$.
  \end{compactenum}
  Let $w''$ denote the first vertex of $P_{ww'}$ contained in $Y_{\delta}$.

>>>>

% \newcommand{\PP}{\mathcal{P}}
\renewcommand{\SS}{\mathcal{S}}

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs.
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results starting with the work of \citet{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}). \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `induced-universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$. This has recently been extended by \citet{EJM}, who show the existence of a `subgraph-universal' graph with $(1+\epsilon)n$ vertices that contains every $n$-vertex planar graph as a subgraph as well as the existence of an `induced universal' graph with $n^{1+o(1)}$ vertices \emph{and} edges.  The former result is the first progress on subgraph-universal graphs for planar graphs since the $O(n^{3/2})$-vertex construction of \citet{babai.chung.ea:on}.
% aking the first progress on this problem since the work of \citet{babai.xx} in 1984.
\end{compactitem}

% \note{DW}{Delete `from 1992' and  `from 2007'? replace ``going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}'' by
	% ``starting with the work of \citet{kannan.naor.ea:implicit}''}
% \note{PM}{Was this a referee request, or just because we don't need to sell it anymore?  I'm fine with the change either way, but didn't implement it.}

@@

All of these results hold for any graph class that has a product structure theorem analogous to \cref{PlanarProduct}; that is, for any graph class  $\mathcal{G}$ where every graph in $\mathcal{G}$ is contained in $H\boxtimes P\boxtimes K_\ell$ where $H$ has bounded treewidth, $P$ is a path, and $\ell$ is bounded.\footnote{It is easily seen that $\tw(H\boxtimes K_\ell) \leq (\tw(H)+1)\ell-1$, so we may assume that $\ell=1$ in this definition.} These applications motivate finding product structure theorems for other graph classes. \citet{DJMMUW20} prove product structure theorems for graphs of bounded Euler genus\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.} and for apex-minor-free graphs\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \defin{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \defin{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \defin{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}, and \citet{DEMWW22} do so for graphs in any minor-closed class and with bounded maximum degree. See \cref{Generalisations} for more precise statements and see \citep{DHJLW21} for a survey on this topic.

@@

\note{PM}{Kind of, but I suggest we keep that paragraph and section anyway. Justification: They're both short and they describe the exact bounds for bounded genus graphs, which you don't get from the one-paragraph description in Section~1.0.}

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarProduct}, we actually prove a substantially stronger result than \cref{kPlanarProduct} using the following definition. A non-empty set $\SS$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

\begin{compactitem}
\item every path in $\SS$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.  A path is \defin{trivial} if it has length 0 and \defin{non-trivial} otherwise.} and
\item for every $v\in V(G)$, the number of paths in $\SS$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem}
Each path $P\in\SS$ is called a \defin{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \defin{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\SS$ for $G$, let $G^{\SS}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\SS$.

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is a subgraph of $G^\SS$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\SS$ for $G$.
\end{obs}

The proof of \cref{AddDummy} is trivial: Given a $k$-plane embedding of a graph $G'$, create a planar graph $G$ by adding a dummy vertex at each crossing point. For each edge $vw\in E(G')$ there is a path $P$ in $G$ between $v$ and $w$ of length at most $k+1$ in which every internal vertex is a dummy vertex. Let $\SS$ be the set of such paths $P$. For each vertex $v$ of $G$, at most two paths in $\SS$ use $v$ as an internal vertex (since no original vertex of $G'$ is an internal vertex of a path in $\SS$). Thus $\SS$ is a $(k+1,2)$-shortcut system for $G$, such that $G'\subseteq G^\SS$. This idea can be pushed further to obtain a rough characterisation of $k$-planar graphs, which is interesting in its own right, and is useful for showing that various classes of graphs are $k$-planar (see \cref{Characterisation}).

@@

\begin{thm}
\label{ShortcutProduct}
Let $G$ be a subgraph of $H\boxtimes P \boxtimes K_\ell$, for some graph $H$ of treewidth at most $t$ and for some path $P$.
Let $\SS$ be a $(k,d)$-shortcut system for $G$. Then $G^\SS$ is a subgraph of $J\boxtimes P\boxtimes K_{d\ell(k^3+3k)}$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$ and some path $P$.
\end{thm}

@@

\note{DW}{Use $\SS$ for shortcut system, okay?}
\note{PM}{Ok.  I've implemented this change.  I was careful, but will still reread.}

@@

\cref{ShortcutProduct} is applicable for many graph classes in addition to $k$-planar graphs. Here is one example. The \defin{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\SS$ for some $(k,2k\Delta^{k})$-shortcut system $\SS$; see \cref{PowerShortcut}. Theorems~\ref{PlanarProduct}(b) and \ref{ShortcutProduct} then imply:

@@

%The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
%
%\begin{thm}
%Let $\SS$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\SS$ is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\SS$ for $G$, $G^\SS\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\SS$ for $G$, $G^\SS \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\note{DW}{delete the above three theorems, they are never used}

@@

The following result is the main technical contribution of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so too does $G^\SS$ for every shortcut system $\SS$ of $G$.

\begin{restatable}{thm}{mmg}
	\label{ShortcutPartition}
	Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\SS$ be a $(k,d)$-shortcut system for $G$.  Then $G^\SS$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$.
\end{restatable}

@@

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, a $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.

@@

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

@@

\begin{proof}
  Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.

@@

\note{DW}{Use $\SS$ for the given shortcut system and $\SS$ for the partition?}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{P}:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\mathcal{P}$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\mathcal{P}$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\SS/\mathcal{P}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

\referee{2}{Page 8.  Can you add a brief explanation why $\mathcal{P}$ is a
	partition of $V(G)$?  It is clear that the sets in $\mathcal{P}$ are
	disjoint, but why do they cover $V(G)$?}
\note{PM}{I added an explanation in a footnote since this is obvious.  In particular, for any function $f$, $\{f^{-1}(x):x\in\mathop{codomain}(f)\}$ is a partition of the $\mathop{domain}(f)$.}

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\mathcal{P}$ has small layered width with respect to the layering $\mathcal{L}$ and that
  \item $J$ has small treewidth.
\end{compactenum}

@@

First we need to understand the relationship between sets $S_x\in\mathcal{P}$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

@@

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\SS$ contains a path $P$, with $v\in V(P)$, that crosses $x$.  The path $P$ contains a proper subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

Next we complete Step~(i) and show that $\mathcal{P}$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

\begin{proof}
  Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v$ participates in $x$.
  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$. Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\SS$ includes $w$ as an internal vertex then $w$ contributes at most one vertex, itself, to $S_x\cap L_i$.

@@

  Otherwise, consider some path $P\in\SS$ that contains $w$ as an internal vertex.  If $w\in L_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap L_i$.  If $w\in L_{i-1}\cup L_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap L_i$. If $w\in L_{i-j}\cup L_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap L_i$.

  For any $j$, the number of vertices $w\in L_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\SS$. Therefore,
  \[  |S_x\cap L_i|\le d\ell  \, \Big(k+1 + 2k + \sum_{j=2}^k 2(k-j)\Big) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +3) \enspace . \qedhere
  \]
\end{proof}

@@

  If $x_\delta\in C_x$, it is because $x_\delta x'\in E(J)$ for some $T$-descendant $x'$ of $x$.  This implies $G^{\SS}$ contains an edge $vw$ with $v\in S_{x'}$ and $w\in S_{x_\delta}=S_\delta$.  This implies that $\SS$ contains a $vw$-shortcut $P_{vw}$.  Let $v'$ be the second-last vertex of $P_{vw}$ (so $v'w\in E(G)$).

  Since $w\in S_{\delta}$, $a(w)=x_\delta$, so $w$ participates in $x_\delta$, so at least one of the following is true:
  \begin{compactenum}
    \item There exists $w'\in V(G)$ such that $\SS$ contains a $ww'$-shortcut $P_{ww'}$ that has an internal vertex in $Y_{\delta}$; or
    \item $w\in Y_\delta$; in this case, define $w':=w$ and let $P_{ww'}$ be the 1-vertex path $w$.
  \end{compactenum}
  Let $w''$ denote the first vertex of $P_{ww'}$ contained in $Y_{\delta}$.
